---
title: "Exercise Evaluation and Prediction Models"
author: "Dirk Mischke"
date: "February 20, 2015"
output: html_document
---
## Executive Summary

The data collected by Velloso et al. is data from 4 different sensors, one belt around the
waist, one on the upper arm, one on the forearm and one on a dumbbell held by the participant.
The sensors recorded the movements of the person during performing the specific weight 
lifting exercise.

The idea is to collect enough data from different ways of doing the exercise to be able to
select a model, with which to predict when the exercise was performed correct and when not.

We will use the provided Training data set with 19622 observations and 160 variables
to try to find the best model to predict the quality of performance ('classe' variable)
for the provided Test data set.

Loading the Training data:
```{r}
trainurl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
download.file(trainurl, destfile="~/Documents/Coursera/Data Science/Practical Machine Learning/Course Project/pml-training.csv", method="curl")
train <- read.csv("pml-training.csv", header=TRUE, sep=",", na.strings=c("NA","#DIV/0!",""))
```
Loadind the Test data:
```{r}
testurl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(testurl, destfile="~/Documents/Coursera/Data Science/Practical Machine Learning/Course Project/pml-testing.csv", method="curl")
test <- read.csv("pml-testing.csv", header=TRUE, sep=",", na.strings=c("NA","#DIV/0!",""))
```
I transformed missing values and '#DIV/0' entries to 'NA' values.

## Initial exploration of the Training data

To find the best model to predict the quality of performance, we have to examine the data
first, to gain some understanding about the available information and about which of the 
present variables should be included in our analysis.

```{r,include=FALSE}
library(dplyr)
library(caret)
library(randomForest)
```

```{r,results='hide'}
str(train)
summary(train)
```

After taking a look at the structure and the summary of the training data set it is obvious that a large number of variables have more than 19200 'NA' entries or missing values. We can safely exclude those from our data set.

## Cleaning the data

Omitting all variables with more than 19200 NA's or missing values:

```{r,comment=NA}
train0 <- select(train, -starts_with("X"))
train0 <- select(train0, -starts_with("raw_time"))
train0 <- select(train0, -starts_with("new_window"))
train0 <- select(train0, -starts_with("kurtosis"))
train0 <- select(train0, -starts_with("skewness"))
train0 <- select(train0, -starts_with("max_"))
train0 <- select(train0, -starts_with("min_"))
train0 <- select(train0, -starts_with("amplitude"))
train0 <- select(train0, -starts_with("var_"))
train0 <- select(train0, -starts_with("avg_"))
train0 <- select(train0, -starts_with("stddev_"))
```

The factor variables 'user_name' and 'cvtd_timestamp' will also be omitted, because
they are just identifiers and also not helpful in predicting the quality of the weight lifting exercise.

```{r,comment=NA}
train0 <- select(train0, -starts_with("user_"))
train0 <- select(train0, -starts_with("cvtd_"))
```
Now we have a data set with 54 variables left for modeling.

## Preprocessing

```{r,comment=NA}
inTrain <- createDataPartition(y=train0$classe, p=0.75, list=FALSE )
train1 <- train0[inTrain,]
test1 <- train0[-inTrain,]
```

To help us choose the right model approach we have to consider that the question at hand
is better described by a classification model and not for example a regression model.
We know that a powerful model is the random Forest method, which we will use for this
case. A positive sideeffect is, that with random Forests we don't need cross validation
to protect from overfitting.
```{r,comment=NA}
modran <- randomForest(classe ~ ., data=train1)
predran1 <- predict(modran, test1, type = "class")
confusionMatrix(predran1, test1$classe)
```

## Conclusion

The overall accuracy of our chosen prediction model is 0.99, which is a very good
value, giving confidence for predicting the actual Test data set. The out-of-sample
error may be higher due to the complexity of the collected data. It is very difficult
to predict movements and specifically if those movements are correct or represent
a specific mistake. This is one of the insights of the original study, that there
is almost an infinite amount of possible ways to do an exercise, which makes it very
challenging to predict.

## Cleaning the Test data set in similar fashion as the Training data set:
```{r,comment=NA}
test0 <- select(test, -starts_with("X"))
test0 <- select(test0, -starts_with("user_"))
test0 <- select(test0, -starts_with("raw_time"))
test0 <- select(test0, -starts_with("cvtd_"))
test0 <- select(test0, -starts_with("new_window"))
test0 <- select(test0, -starts_with("kurtosis"))
test0 <- select(test0, -starts_with("skewness"))
test0 <- select(test0, -starts_with("max_"))
test0 <- select(test0, -starts_with("min_"))
test0 <- select(test0, -starts_with("amplitude"))
test0 <- select(test0, -starts_with("var_"))
test0 <- select(test0, -starts_with("avg_"))
test0 <- select(test0, -starts_with("stddev_"))
```

## Creating submission files, applying the prediction model chosen from the analysis of the training data 

```{r}
predran0 <- predict(modran, test, type = "class")

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predran0)
```

Source:

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.

Read more: http://groupware.les.inf.puc-rio.br/har#weight_lifting_exercises#ixzz3SKam4z8q

